Refinement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-20 rigid_bodies.tex
Further revision of the Discretization Effect section

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state of the system immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
This justifies its more widespread use.
However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:modified_h} and Appendix \ref{sec:rigid body shadow hamiltonian}.

@@

\subsection{Backward Error Analysis}
\label{sec:modified_h}

@@

In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for the liquid water system using several NVT-MD methods with different time step sizes.
All simulations were carried out by specifying $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
the mean potential energy per molecule $\langle U/N \rangle$,
the mean internal virial per molecule $\langle W/N \rangle$, and
the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
The horizontal, dotted lines are meant to help visualizing the departure of the computed ensemble averages from their expected values.
They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods.
Error bars were obtained using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.

@@

Fig.~\ref{fig:properties} contains results obtained by using five different NVT integrators.
First, for the sake of comparison, we test our measure-preserving extension to rigid bodies of the method of Martyna \textit{et al}. \cite{Martyna_1996} without doing any refinement, such as detailed in Sec.~\ref{sec:nvt}.
Another method we use with no refinement is that of Kamberaj \textit{et al}. \cite{Kamberaj_2005}.
We will show that, besides the spurious energy flux between the thermostats, such method exhibits some discretization-related issues....................................

We have included the results of applying our method, but using the approximate solution of rotation due to Miller \textit{et al}. \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
Recall that the computed averages for both refined schemes are obtained via reweighting by employing Eq.~\eqref{eq:reweighting}.

Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005}, because that is not the only problematic issue.
Particularly, from Fig.~\ref{fig:properties}(a) we note that the simulated temperature decreases systematically with $\timestep$.
This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al}. \cite{Martyna_1996}. Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of $2~\text{fs}$.
Thus, the results for the method of Martyna \textit{et al}. \cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.

>>>>

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state of the system immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
This justifies its more widespread use.
However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.

@@

\subsection{Backward Error Analysis}
\label{sec:backward error analysis}

@@

In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for the liquid water system using several NVT-MD methods with different time step sizes.
All simulations were carried out by specifying $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
the mean potential energy per molecule $\langle U/N \rangle$,
the mean internal virial per molecule $\langle W/N \rangle$, and
the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
The horizontal, dotted lines are meant to help visualizing the departure of the computed ensemble averages from their expected values.
They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods.
Error bars represent 95\% confidence intervals obtained using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.

@@

Fig.~\ref{fig:properties} contains results obtained by using five different NVT integrators.
Our measure-preserving extension to rigid bodies of the splitting scheme developed by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
First, the method is applied in a conventional (that is, refinement-free) way using the unsplit solution of Ref.~\citenum{Silveira_2017} for handling rotations.
Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
Next, we substitute the unsplit rotation by the simpler NO-SQUISH splitting of Miller \textit{et al}. \cite{Miller_2002}, denoting the obtained integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
We note that our refinement procedure is rigorously sustained (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
However, NO-SQUISH rotations have been shown to approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
The fourth integrator we test here is a refined version of the velocity-rescaling method of Bussi \textit{et al}. \cite{Bussi_2007}, which differs from the other ones for involving a single stochastic (rather than deterministic) thermostat.
This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}. \cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux between thermostats described in the preceding section.
This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.

Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated by simple averages taken from the sampled configurations in the case of a conventional method (either Martyna or Kamberaj), whereas reweighting via Eq.~\eqref{eq:reweighting} is carried out in the case of a refined one.

%Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005}, because that is not the only problematic issue.
Particularly, from Fig.~\ref{fig:properties}(a) we note that the simulated temperature decreases systematically with $\timestep$.
This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al}. \cite{Martyna_1996}. Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of $2~\text{fs}$.
Thus, the results for the method of Martyna \textit{et al}. \cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
